Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-11-22 ms/review-responses.tex

\reply{
    This is a natural question that we asked ourselves while developing the simulation using the built-in visualization tools of the SLiM GUI. We have now added a supplementary figure (Figure \ref{fig:wm_v_dd}) showing the distribution of individuals in our density-dependant spatial model and a continuous-space Wright-Fisher simulation without density dependance. 
    We encourage the reviewer to run our SLiM recipe using the GUI at different parameter values
    to see for themselves how quickly equilibrium is attained --
    we are tempted to include movies of this as supplementary material,
    but that seems excessive.
    Also note that we did \emph{not} include competitive interaction to avoid clumping --
    we included it to make the model biologically realistic
    (so that, for instance, the population does not grow without bound),
    and lack of clumping is a \emph{consequence} of this choice.
    We have clarified this in the text. \revref
}

\begin{point}{}
Second, the demographic process used here involves birth and death of individuals. Does the population survive asymptotically or, like any birth-death process, eventually dies with probability one? In fact, one needs to know a little about the dynamics of the population size to decide whether the corresponding process is reasonable from a biological standpoint.
\end{point}

\reply{
    As in the last point, as part of sanity checking our models
    we monitored the population size over time while with SLiM.
    All simulations maintained asymptotic populations, and none of the runs we started ended because populations crashed.
    We agree that the population model will die out eventually with probability 1
    (it is a Markov process with a single accessible absorbing state (death),
    and since it isn't a branching process, doesn't grow without bound).
    However, it will take an extremely long time to hit that state,
    so that extinction is not a practical consideration.
    We have added a clarifying note about this. \revref
}

@@

\begin{point}{}
    All in all, more efforts need to be made here in my opinion to show that the forward-in-time simulations generate sensible outcomes. Sensible in terms of the behavior of the population demography at equilibrium (provided such equilibrium indeed exists) along with that of the spatial distribution of individuals. The authors could provide some guarantee of the good behavior of their model as evidenced from simulations using a broad range of parameter values for generating data. Alternatively, they could elect to use the spatial-Lambda-Fleming-Viot model for their simulations, which, in my opinion would seem the most sensible option given that (1) it is possible to run backward-in-time simulations under this model, thereby saving a lot of computation time and (2) it is a well-studied model with good mathematical and biological properties and (3) it is implemented in a publicly available software program (\url{https://github.com/jeromekelleher/discsim}).
\end{point}

\reply{
    Hopefully, the additional plots 
    and discussion around demographic modeling (notably, the discussion at \llname{demog_disco})
    helps to increase the transparency in this model.
    We agree that demographic realism is a very important consideration,
    and due to the often hard-to-intuit nature of spatial demography,
    simulations need to be carefully sanity checked.
    (Furthermore, what is ``realistic'' for one species will not be for another!)
    We have added an Appendix comparing our model to a stepping-stone \llname{apx:stepping_stone}, 
    because we think these are the most familiar and widely used class of spatial models.
    We find that many features of our model are well approximated by stepping-stone models, and that for statistics like $\theta_{\pi}$ the stepping stone model results approach our continuous space model as the number of demes used to describe the landscape increases. 
}

    We definately think that a comparison to the spatial Lambda-Fleming-Viot
    would be an extremely useful thing to do, for the reasons the reviewer mentions,
    but this is beyond the scope of the current study.
    The SLFV model is only known to be an approximation of one particular biologically explicit model
    (of patchy extinction-recolonization)
    and is only conjectured to be a good approximation
    for spatial models more generally.
    To reiterate, the definition of the SLFV was explicitly motivated by a desire to approximate
    spatial models, especially Bolker-Pacala models like ours.
    Therefore, our study does important preliminary work
    for simulation study of the spatial Lambda-Fleming-Viot,
    since it works through some of the issues in simulating biologically realistic spatial models,
    and describes some of the patterns.

@@

\begin{point}{\autoref{fig:genparams}:}
    I do not understand why the neighborhood size varies to the same extent in the random mating model as it does for the spatial model. For the random mating model, I would have expected the neighborhood size to be equal to the census size since all individuals have the same probability of being a parent of any given offspring. From [the paragraph at \revref], it is clear that the spatial model would converge to the random mating model when the mean parent-offspring distance tends to infinity only if we were to ignore the impact of range edges. I am thus wondering whether the variation of neighborhood size one observes in Fig 2 for the random mating model is just a consequence of border effects. If that is the case, then the authors should state it clearly and try to justify it from a biological perspective.
\end{point}

\reply{
    It's a good point that neighborhood size, as defined to be ``number of possible parents''
    should indeed be the census size.
    However, we are using $N_W$ defined in the same way for both spatial and random mating simulations
    as a way of comparing spatial to ``nonspatial'' models
    keeping other aspects of demography (almost) the same.
    We've added a note clarifying this, at \revref and \llname{randomNW}.
    This is indeed part of what's going on.
    We have added some discussion of the census size scaling \llname{censusscaling},
    but haven't added a detailed discussion of whether or not most species experience
    decreased habitat quality at range edges,
    since that seems like it is getting too far into the weeds.
}

### 2019-11-22 ms/spaceness.tex
added a sentence to the demog model section for clarity
editz

Our demographic model includes parameters that control population density ($K$), \revpoint{1}{2}
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantitites to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations. 
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citetext{\url{https://github.com/tskit-dev/pyslim}}.

@@

There are a few additional things to note about Figure~\ref{fig:genparams}. 
First, all three quantities are non-monotone with neighborhood size. 
Census size largely declines as neighborhood size increases for both the spatial and random mating models. 
However, for spatial models this decline only begins for neighborhood size $\geq 10$. 
Spatial and random mating models are indistinguishable from one another
for neighborhood sizes larger than 100.
Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000. The scaling of census sizes in both random-mating and spatial models appears to be related to two consequences of the spatial competition function: the decline of fitness at range edges, which effectively reduces the habitable area by one $\sigma$ around the edge of the map and so results in a smaller habitable area at high $\sigma$ values; and variation in the equilibrium population density given varying competition radii.
Furthermore, census size increases in spatial models as neighborhood size increases from 2 to 10.
This may reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.  \llabel{censusscaling}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary_loglog_qqlines.pdf}
    \caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS. Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors. In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype. As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}). Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding the first 10 PC coordinates as covariates (\textbf{C}). Quantile-quantile plots in (\textbf{D}) show inflation of $-\log_{10}(p)$ after PC correction for simulations with spatially structure environments, with line colors scaled to the neighborhood size of each simulation.
}
\label{fig:gwas}
\end{figure*}

@@

\subsection{Where are natural populations on this spectrum?}
For how much of the tree of life do spatial patterns circumscribe genomic variation? In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms to get an idea of how likely dispersal is to play an important role in patterns of variation. These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$. \llabel{popsizecaveat}
In addition, these empirical examples are likely biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density). However, from the available data we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa. At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants. Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference.
These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citep{Rousset1997} range from 40 to 410 depending on the method of estimation. \revpoint{3}{7}
Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered. Surprisingly, even some flying insects with huge census population sizes fall in this group, including fruit flies (\textit{D.~melanogaster}) and mosquitoes (\textit{A. aegypti}). Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in northeastern Europe \citep{Ringbauer2017}. For these species demographic inference and summary statistics are likely to reflect minimal bias from spatial effects as long as dispersal is truly continuous across the landscape. While that is so we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Demographic model}
\llabel{apx:demographic_model}

We chose our demographic model so that every individual has on average $1/L$ offspring
each time step,
and if the local population density of an individual is $n$,
then their probability of survival until the next time step is
(equation \eqref{eqn:p_survival}):
\begin{equation}
    p = \min\left( 0.95, \frac{1}{1 + n / (K (1+L))} \right) .
\end{equation}
We capped survival at 0.95 so that we would not have exceptionally long-lived individuals
in sparsely populated areas --
otherwise, an isolated individual might live for a very long time.
Since $1 - p \approx n / (K (1 + L))$,
mortality goes up roughly linearly with the number of neighbors (on a scale given by $K$),
as would be obtained if, for instance, mortality is due to agonistic interactions.
Ignoring migration,
a region is at demographic equilibrium if the per-capita probability of death
is equal to the birth rate,
i.e., if $1 - p = 1/L$.
(Note that there is no effect of age in the model,
which would make the analysis more complicated.)
Solving this for $n$,
we get that in a well-mixed population, the equilibrium density should be around
\begin{equation}
    n = K \frac{L+1}{L-1}
\end{equation}
individuals per unit area.
At this density, the per-capita death rate is $1/L$, so the mean lifetime is $L$.
This equilibrium density is \emph{not} $K$, but (since $L=4$) is two-thirds larger.
However, in practice this model leads to a total population size which is around $K$
multiplied by total geographic area
(but which depends on $\sigma$, as discussed above).
The main reason for this is that
since offspring tend to be near their parents, individuals tend to be ``clumped'',
and so experience a higher average density than the ``density'' one would compute
by dividing census size by geographic area \citep{lloyd1967crowding}.
To maintain a constant expected total population size
would require making (say) $K$ depend on $\sigma$;
however, typical local population densities might then be more dissimilar.

>>>>

Our demographic model includes parameters that control population density ($K$), \revpoint{1}{2}
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantitites to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations,
and used mean population density ($\rho$, census size divided by total area)
to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$. \revpoint{3}{6}
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citetext{\url{https://github.com/tskit-dev/pyslim}}.

@@

There are a few additional things to note about Figure~\ref{fig:genparams}. 
First, all three quantities are non-monotone with neighborhood size. 
Census size largely declines as neighborhood size increases for both the spatial and random mating models. 
However, for spatial models this decline only begins for neighborhood size $\geq 10$. 
Spatial and random mating models are indistinguishable from one another
for neighborhood sizes larger than 100.
Census sizes range from around 14,000 at low $\sigma$ in the random mating model to 10,000 for both models when neighborhood sizes approach 1,000.
The scaling of census sizes in both random-mating and spatial models appears to be related to two consequences of the spatial competition function: the decline of fitness at range edges, which effectively reduces the habitable area by one $\sigma$ around the edge of the map and so results in a smaller habitable area at high $\sigma$ values; and variation in the equilibrium population density given varying competition radii.
Furthermore, lack of available mates contributes to lower census sizes at small $N_W$
for the spatial model
(We calculate $N_W$ for both in the same way,
even though the number of ``available mates'' is much larger with random mating). \llabel{randomNW}
Furthermore, census size increases in spatial models as neighborhood size increases from 2 to 10.
This may reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.  \llabel{censusscaling}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary_loglog_qqlines.pdf}
    \caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS. Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors. In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype. As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}).
    Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding the first 10 PC coordinates as covariates (\textbf{C}).
    Quantile-quantile plots in (\textbf{D}) show inflation of $-\log_{10}(p)$ after PC correction for simulations with spatially structured environments, with line colors showing the neighborhood size of each simulation.
}
\label{fig:gwas}
\end{figure*}

@@

\subsection{Where are natural populations on this spectrum?}
For how much of the tree of life do spatial patterns circumscribe genomic variation?
In Table \ref{table:NStable} we gathered estimates of neighborhood size from a range of organisms
to get an idea of how strongly local geographic dispersal affects patterns of variation.
This is an imperfect measure:
some aspects of genetic variation are most strongly determined by neighborhood size \citep{Wright1946},
others (e.g., number of segregating sites)
are more strongly determined by global $N_e$ or by the ratio of the two.
% These values should be compared to our simulation results with some caution, as though we expect neighborhood size to have similar effects across species of varying global $N_e$ \citep{Wright1946}, in our study we evaluated only a relatively small population of $\approx 10,000$. 
\llabel{popsizecaveat}
In addition, these empirical examples are likely biased towards small-neighborhood species (because few studies have quantified neighborhood size in species with very high dispersal or population density). However, from the available data we find that neighborhood sizes in the range we simulated are fairly common across a range of taxa. At the extreme low end of empirical neighborhood size estimates we see some flowering plants, large mammals, and colonial insects like ants. Species such as this have neighborhood size estimates small enough that spatial processes are likely to strongly influence inference.
These include some human populations such as the Gainj- and Kalam-speaking people of Papua New Guinea, in which the estimated neighborhood sizes in \citep{Rousset1997} range from 40 to 410 depending on the method of estimation. \revpoint{3}{7}
Many more species occur in a middle range of neighborhood sizes between 100 and 1000 -- a range in which spatial processes play a minor role in our analyses under random spatial sampling but are important when sampling of individuals in space is clustered. Surprisingly, even some flying insects with huge census population sizes fall in this group, including fruit flies (\textit{D.~melanogaster}) and mosquitoes (\textit{A. aegypti}). Last, many species likely have neighborhood sizes much larger than we simulated, including modern humans in northeastern Europe \citep{Ringbauer2017}. For these species demographic inference and summary statistics are likely to reflect minimal bias from spatial effects as long as dispersal is truly continuous across the landscape. While that is so we caution that association studies in which the effects of population structure are confounded with spatial variation in the environment are still sensitive to dispersal even at these large neighborhood sizes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Demographic model}
\llabel{apx:demographic_model}
Local population regulation is controlled by two parameters, $L$, and $K$.
Here, we show that these should be close to the average lifespan of an individual
and the average number of individuals per unit area, respectively.
We chose our demographic model so that every individual has on average $1/L$ offspring
each time step,
and if the local population density of an individual is $n$,
then their probability of survival until the next time step is
(equation \eqref{eqn:p_survival}):
\begin{equation}
    p = \min\left( 0.95, \frac{1}{1 + n / (K (1+L))} \right) .
\end{equation}
We capped survival at 0.95 so that we would not have exceptionally long-lived individuals
in sparsely populated areas --
otherwise, an isolated individual might live for a very long time.
Since $1 - p \approx n / (K (1 + L))$,
mortality goes up roughly linearly with the number of neighbors (on a scale given by $K$),
as would be obtained if, for instance, mortality is due to agonistic interactions.
Ignoring migration,
a region is at demographic equilibrium if the per-capita probability of death
is equal to the birth rate,
i.e., if $1 - p = 1/L$.
(Note that there is no effect of age in the model,
which would make the analysis more complicated.)
Solving this for $n$,
we get that in a well-mixed population, the equilibrium density should be around
\begin{equation}
    n = K \frac{L+1}{L-1}
\end{equation}
individuals per unit area.
At this density, the per-capita death rate is $1/L$, so the mean lifetime is $L$.
This equilibrium density is \emph{not} $K$, but (since $L=4$) is two-thirds larger.
However, in practice this model leads to a total population size which is around $K$
multiplied by total geographic area
(but which depends on $\sigma$, as discussed above).
The main reason for this is that
since offspring tend to be near their parents, individuals tend to be ``clumped'',
and so experience a higher average density than the ``density'' one would compute
by dividing census size by geographic area \citep{lloyd1967crowding}.
To maintain a constant expected total population size
would require making (say) $K$ depend on $\sigma$;
however, typical local population densities might then be more dissimilar.
